---
title: "Auditing in-vivo transferability of a transcriptomic signature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing in-vivo transferability of a transcriptomic signature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`transferaudit` asks whether a transcriptomic signature localises to its expected
cell compartment in a single-cell reference atlas **more than a size-matched
random gene set would**. This vignette uses a small simulated atlas so it runs
anywhere, and shows the central contrast the package was built to expose:

* a **prior-knowledge / identity** signature *localises* to its compartment;
* a **culture / off-target** signature does *not*, even if both are "real" gene sets.

```{r}
library(transferaudit)
set.seed(1337)

## ---- simulate an atlas: 1500 cells across 5 compartments, 1200 genes --------
comps   <- c("Mesenchymal_osteo", "Myeloid", "Tcell", "Endothelial", "Epithelial")
ncell   <- 1500
compart <- sample(comps, ncell, replace = TRUE)
genes   <- paste0("G", seq_len(1200))

# baseline log-normal expression
expr <- matrix(rpois(length(genes) * ncell, lambda = 1), nrow = length(genes),
               dimnames = list(genes, NULL))
expr <- log1p(expr)

# define two ground-truth programs of 40 genes each
osteo_id   <- genes[1:40]      # marks the Mesenchymal_osteo compartment (identity)
myeloid_pr <- genes[41:80]     # marks the Myeloid compartment (off-target for osteo)

# inject identity signal: osteo_id high in osteoprogenitors, myeloid_pr high in myeloid
expr[osteo_id,  compart == "Mesenchymal_osteo"] <-
  expr[osteo_id,  compart == "Mesenchymal_osteo"] + 2
expr[myeloid_pr, compart == "Myeloid"] <-
  expr[myeloid_pr, compart == "Myeloid"] + 2
```

Build the genes × compartments matrix once:

```{r}
M <- build_compartment_matrix(expr, compart)
dim(M)            # genes x 5 compartments
```

Score each program against the **osteoprogenitor** compartment:

```{r}
osteo_res <- transferability_score(M, up = osteo_id,   target = "Mesenchymal_osteo", n_perm = 1000)
myelo_res <- transferability_score(M, up = myeloid_pr, target = "Mesenchymal_osteo", n_perm = 1000)

c(osteo_transfer_z   = round(osteo_res$transferability_z, 2),
  osteo_p_more       = signif(osteo_res$p_localizes_more, 3),
  myeloid_transfer_z = round(myelo_res$transferability_z, 2),
  myeloid_p_more     = signif(myelo_res$p_localizes_more, 3))
```

The osteoprogenitor-identity program sits far in the right tail of the random null
(high transferability z, small p): it *transfers*. The myeloid program — equally
"real", but marking a different compartment — does not localise to osteoprogenitors.

Audit a whole panel in one call:

```{r}
panel <- list(
  osteo_identity = list(up = osteo_id),
  myeloid_offtgt = list(up = myeloid_pr),
  random_set     = list(up = sample(genes, 40))
)
audit_panel(M, panel, target = "Mesenchymal_osteo", n_perm = 1000)
```

`transfers == TRUE` requires both that the target compartment ranks first and that
`p_localizes_more < 0.05`. This is exactly the calibrated test used in the
accompanying study to show that data-driven in-vitro osteogenic signatures fail to
localise to in-vivo osteoprogenitors while prior-knowledge osteogenic gene sets do.
